Experience Report: Haskell in Computational Biology

### 2012-07-01 mrfy_experience_report.tex
performance section soundly whacked

\begin{document}
%\parskip=0.9\parskip % cheating
%\advance \parskip by 0pt plus 0.1pt

@@

We~also found an opportunity to improve performance through parallelism.
Calls on query sequences between beta strands are independent,
as are calls on different placements.
We~found them easy to parallelize
using
\texttt{Control.Parallel}:
we substituted
\texttt{parmap rseq} for \texttt{map}.
On a 48-core, 2.3GHz AMD Opteron 6176 system, 
parallel performance is good up to about 12~cores.
\figref{fig:efficiency} shows parallel efficiency when
using a model of an ``8-bladed beta propeller;''
the model has 343 ordinary nodes and 40~beta strands TOTALLING HOW
MANY NODES?
We~benchmarked \mrfy\ using a query sequence of
592 amino acids.
Figure~\ref{fig:speedup} SHOWS NOTHING OF VALUE BECAUSE RANDOMNESS
BREAKS EVERYTHING.  AND IT SHOULD HAVE ERROR BARS.
(Parallel efficiency is 1.0 when the speedup is equal to the number of cores
used.
Lower speedups mean lower efficiency.)

@@

MRFy's implementation of Viterbi's algorithm appears dramatically
slower than the C++ version implemented in \mrfy\'s predecessor,
SMURF.
Both \mrfy\ and SMURF, when provided with a Markov random field containing no beta strands,
simply compute the optimal solution using Viterbi's algorithm.
For both \mrfy\ and SMURF, this task admits no parallelism.
On a 2.6GHz Intel Core i7,
given a micro-benchmark consisting of a contrived query sequence
of 2000 amino acids (which represents a large protein),
and a hidden Markov model of 343 nodes and no beta strands,
\mrfy\ takes 2.32 seconds to compute a Viterbi solution.
SMURF performs the same computation in 0.29 seconds.

But unlike SMURF, \mrfy\ doesn't have to run on large models.
For example, on the ``8-bladed propeller'' benchmark, the beta strands
break the Markov random field into 41 pieces, each of which is much smaller
than the original (typically 10 or fewer nodes rather than 343).
And, each \texttt{placement} breaks the query sequence into 41 pieces, each of
which is much smaller than the original (typically 20 or fewer amino acids
rather than 592).
Thus, \mrfy\ repeatedly runs Viterbi's algorithm on 
a~great many smaller models, and the independence of these models
means that \mrfy\ can execute them in parallel.
Here, performance is good enough that \mrfy\ can do its job.
Given a Markov random field with a 12-stranded ``beta sandwich''
structure,
\mrfy\ typically computes an alignment within a minute;
SMURF allocates over 16GB of memory
and does not terminate even after eight hours.

Clearly, micro-benchmarks do not tell the whole story.
Our goal is to solve a biological problem, and though our implementation of
Viterbi's algorithm is comparatively slow, we can still solve a biological problem
quickly.
%%% Had we implemented the SMURF algorithm in Haskell, we believe that performance
%%% would be unacceptable.
%%% However, for the algorithm underlying MRFy, we are satisfied with the runtime
%%% performance of Haskell and GHC.
%
% Nonetheless, we expect that we can still improve MRFy's run-time performance
% with further profiling and tuning.  
% And we expect that if we cross the street, we will get to the other side.

>>>>

\begin{document}
\parskip=0.9\parskip % cheating
\advance \parskip by 0pt plus 0.1pt

@@

Even with this trick,
MRFy's implementation of Viterbi's algorithm is much
slower than the C++ version implemented in \mrfy\'s predecessor,
SMURF.
For example, on a micro-benchmark that searches for a
structural motif of 343 nodes in a protein of 2000 amino acids,
\mrfy\ takes 2.32~seconds and
SMURF takes 0.29~seconds.

But \mrfy's job is not to run Viterbi's algorithm on large models;
\mrfy's job is to detect homologies to structures on which Viterbi's
algorithm is intractable.
We~benchmarked \mrfy\ 
using a model of an ``8-bladed beta propeller,''
which has 343 nodes, of which NNN appear in 40~beta strands.
The models between beta strands typically have at most 10~nodes.
We~used a query sequence of 592 amino acids, but each placement breaks
the sequence into 41~pieces, each of which typically has at most 20 amino
acids.
CAN WE PLEASE REPLACE THE TEXT ABOVE WITH A DESCRIPTION OF WHAT'S
GOING ON IN THE CODE WE ACTUALLY RAN???
On~these smaller problems, 
performance is good enough that \mrfy\ can do its job.
%%
%%
%%
SUDDEN CHANGE OF BENCHMARK CONSIDERED HARMFUL.
%%
%%
Given a Markov random field with a 12-stranded ``beta sandwich''
structure,
\mrfy\ typically computes an alignment within a minute;
SMURF allocates over 16GB of memory
and does not terminate even after eight hours.

Because \mrfy\ works on many subproblems independently, we~found it easy
to improve performance through parallelism.
Calls on query sequences between beta strands are independent,
as are calls on different placements.
Using
\texttt{Control.Parallel}, parallelizing this computation was as easy
as  substituting
\texttt{parmap rseq} for \texttt{map}.
On a 48-core, 2.3GHz AMD Opteron 6176 system, 
parallel performance is good up to about 12~cores.
\figref{fig:efficiency} shows parallel efficiency when
using a model of an ``8-bladed beta propeller;''
the model has 343 ordinary nodes and 40~beta strands TOTALLING HOW
MANY NODES?
We~benchmarked \mrfy\ using a query sequence of
592 amino acids.
Figure~\ref{fig:speedup} SHOWS NOTHING OF VALUE BECAUSE RANDOMNESS
BREAKS EVERYTHING.  AND IT SHOULD HAVE ERROR BARS.
(Parallel efficiency is 1.0 when the speedup is equal to the number of cores
used.
Lower speedups mean lower efficiency.)

@@

%%% Had we implemented the SMURF algorithm in Haskell, we believe that performance
%%% would be unacceptable.
%%% However, for the algorithm underlying MRFy, we are satisfied with the runtime
%%% performance of Haskell and GHC.
%
% Nonetheless, we expect that we can still improve MRFy's run-time performance
% with further profiling and tuning.  
% And we expect that if we cross the street, we will get to the other side.
